\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-26 paper.qmd
Updated intro section 7.1

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. NLDR methods are commonly used to display clusters, and help to verify the results. For example, @chen2023 illustrates the use of UMAP to identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k). @fig-NLDR-variety (a) is the reproduction of the published plot. It shows three well-separated clusters with big separations.

@@

To determine whether the UMAP representation with the specific (hyper-)parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data. The @fig-umap-author shows three well-separated clusters. However, as shown in @fig-pbmc1-sc, there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-umap-author) does not accurately represent the structure(s) present in PBMC3k dataset. But, when visualizing the model-in-th-data-space some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-pbmc1-sc).

@@

```{r}
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-height: 10
#| fig-width: 10
#| fig-cap: "Model in \\gD{}, on the UMAP layout, and three views of the fit in projections from $9\\text{-}D$, for the PBMC3k data ($(s_1, \ s_2) = (-0.050, \ -0.041)$, $b = 870 \  (30, \ 29)$, $m = 135$, benchmark value to remove large edges is $0.099$) (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/VqqWuE0Jj6A>)."

>>>>

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. NLDR methods are commonly used to display clusters, and help to verify the results. For example, @chen2023 illustrates the use of UMAP to identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k). There are $2622$ single cells. First 9 principal components are used to generate the UMAP. @fig-NLDR-variety (a) is the reproduction of the published plot.

@@

To determine whether the UMAP representation with the specific (hyper-)parameter choice suggested by @chen2023 preserves the data structure present in $9\text{-}D$ (@fig-model-pbmc-author-proj (b)), we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data. The @fig-NLDR-variety (a) shows three well-separated clusters with big separations. However, as shown in @fig-model-pbmc-author-proj, there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-NLDR-variety (a)) does not accurately represent the structure of PBMC3k dataset in $9\text{-}D$. But, when visualizing the *model-in-th-data-space* some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-model-pbmc-author-proj).

@@

```{r}
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-height: 10
#| fig-width: 10
#| fig-cap: "Model in \\gD{}, on the UMAP layout, and three views of the fit in projections from $9\\text{-}D$, for the PBMC3k data ($b = 870 \  (30, \ 29)$, $m = 135$, benchmark value to remove large edges is $0.099$) The full video is available at <https://youtu.be/VqqWuE0Jj6A>."

### 2024-08-26 paper.tex
Updated intro section 7.1

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR methods are commonly used to display clusters, and help to verify
the results. For example, \citet{chen2023} illustrates the use of UMAP
to identify clusters in Human Peripheral Blood Mononuclear Cells
(PBMC3k). Figure~\ref{fig-NLDR-variety} (a) is the reproduction of the
published plot. It shows three well-separated clusters with big
separations.

To determine whether the UMAP representation with the specific
(hyper-)parameter choice suggested by \citet{chen2023} preserves the
original data structure, we visualize the model constructed with UMAP
overlaid on the \(9\text{-}D\) data. The \textbf{?@fig-umap-author}
shows three well-separated clusters. However, as shown in
\textbf{?@fig-pbmc1-sc}, there is no big separation between three
clusters in \(9\text{-}D\). Therefore, the suggested UMAP representation
(\textbf{?@fig-umap-author}) does not accurately represent the
structure(s) present in PBMC3k dataset. But, when visualizing the
model-in-th-data-space some unobserved structures can be seen. Some
clusters have non-linear continuity patterns and high-density patches
(\textbf{?@fig-pbmc1-sc}).

@@

\caption{\label{fig-model-pbmc-author-proj}Model in \gD{}, on the UMAP
layout, and three views of the fit in projections from \(9\text{-}D\),
for the PBMC3k data (\((s_1,  s_2) = (-0.050,  -0.041)\),
\(b = 870   (30,  29)\), \(m = 135\), benchmark value to remove large
edges is \(0.099\)) (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/VqqWuE0Jj6A}).}

>>>>

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR methods are commonly used to display clusters, and help to verify
the results. For example, \citet{chen2023} illustrates the use of UMAP
to identify clusters in Human Peripheral Blood Mononuclear Cells
(PBMC3k). There are \(2622\) single cells. First 9 principal components
are used to generate the UMAP. Figure~\ref{fig-NLDR-variety} (a) is the
reproduction of the published plot.

To determine whether the UMAP representation with the specific
(hyper-)parameter choice suggested by \citet{chen2023} preserves the
data structure present in \(9\text{-}D\)
(Figure~\ref{fig-model-pbmc-author-proj} (b)), we visualize the model
constructed with UMAP overlaid on the \(9\text{-}D\) data. The
Figure~\ref{fig-NLDR-variety} (a) shows three well-separated clusters
with big separations. However, as shown in
Figure~\ref{fig-model-pbmc-author-proj}, there is no big separation
between three clusters in \(9\text{-}D\). Therefore, the suggested UMAP
representation (Figure~\ref{fig-NLDR-variety} (a)) does not accurately
represent the structure of PBMC3k dataset in \(9\text{-}D\). But, when
visualizing the \emph{model-in-th-data-space} some unobserved structures
can be seen. Some clusters have non-linear continuity patterns and
high-density patches (Figure~\ref{fig-model-pbmc-author-proj}).

@@

\caption{\label{fig-model-pbmc-author-proj}Model in \gD{}, on the UMAP
layout, and three views of the fit in projections from \(9\text{-}D\),
for the PBMC3k data (\(b = 870   (30,  29)\), \(m = 135\), benchmark
value to remove large edges is \(0.099\)) The full video is available at
\url{https://youtu.be/VqqWuE0Jj6A}.}
